---
title: "The bootstrap method - Solutions"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data

For this lab, we will use the dataset [sphagnum_cover.csv](sphagnum_cover.csv), from the paper:

> Maanavilja, L., Kangas, L., Mehtätalo, L. and Tuittila, E.‐S. (2015), Rewetting of drained boreal spruce swamp forests results in rapid recovery of Sphagnum production. J Appl Ecol, 52: 1355-1363. [doi:10.1111/1365-2664.12474](https://doi.org/10.1111/1365-2664.12474))

This dataset contains measurements of the percentage cover of sphagnum moss (*sphcover*) for 36 boreal swamps divided into three types (*habitat*): *Dr* = drained, *Re* = rewetted et *Un* = undrained.

```{r}
cover <- read.csv("sphagnum_cover.csv")
str(cover)
```


## 1. Estimation of mean cover for drained swamps

a) From the dataset, extract the *sphcover* values for drained swamps. Calculate the mean percentage cover and its standard error from the usual formula (based on standard deviation and sample size). Finally, calculate the 95% confidence interval based on the $t$ distribution:

$$(\bar{x} + t_{(n-1)0.025} s_{\bar{x}}, \bar{x} + t_{(n-1)0.975} s_{\bar{x}})$$

*Reminder*: The function `qt(p, df)` gives the quantile corresponding to a given cumulative probability `p`, for a $t$ distribution with `df` degrees of freedom.

**Solution**

```{r}
cov_dr <- cover$sphcover[cover$habitat == "Dr"]

n <- length(cov_dr)

# Mean
moy <- mean(cov_dr)
moy

# Standard error
err_type <- sd(cov_dr) / sqrt(n)
err_type

# Confidence interval
ic <- c(moy + qt(0.025, n - 1) * err_type, 
        moy + qt(0.975, n - 1) * err_type)
ic
```


b) Simulate 10,000 bootstrap samples for the mean calculated in a). What is its standard error according to the bootstrap? Does this statistic appear biased? 

**Solution**

```{r}
library(boot)
set.seed(5612)

calc_moy <- function(x, i) mean(x[i])

boot_moy <- boot(cov_dr, calc_moy, R = 10000)

# Erreur-type
sd(boot_moy$t)

# Biais
mean(boot_moy$t) - boot_moy$t0
```

The bias is quite small (compared to the standard error) and could be due to a numerical error rather than a real bias in the statistic.


c) How does the bootstrap distribution differ from a normal distribution? To answer this question, it may be useful to draw a quantile-quantile plot (in the code below, `res` is the result of the bootstrap):

```{r, eval = FALSE}
qqnorm(res$t)
qqline(res$t)
```

**Solution**

```{r}
qqnorm(boot_moy$t)
qqline(boot_moy$t)
```

The bootstrap distribution is skewed: the smallest values are above the line, therefore closer to the mean compared to a normal distribution, while the largest values are also above the line, therefore further from the mean.


d) Calculate the 95% confidence interval of the mean using the BCa method. How does it differ from the one calculated in a) using the classical formula? Can you explain this difference based on the result in c)?

**Solution**

```{r}
boot.ci(boot_moy)
```

The BCa interval (3.43, 13.28) differs from that in a) (1.36, 12.97) especially at the lower bound, and both bounds are corrected upwards. In c), we have seen that the smallest values of the bootstrap distribution are closer to the mean with respect to the normal distribution, whereas the largest values are farther away from it.


## 2. Estimation of differences between habitats

a) Here is the distribution of *sphcover* values in each habitat type.

```{r, message = FALSE}
library(ggplot2)
ggplot(cover, aes(x = habitat, y = sphcover)) +
    geom_boxplot()
```

What are the assumptions of a classical ANOVA model that would describe sphagnum moss cover by habitat type? Do these assumptions appear to be met here?

**Solution**

The ANOVA assumes that the residuals follow a normal distribution of the same variance for each group. Here, the distributions are skewed and the variance seems smaller for the *Dr* group seems than the others.


b) Fit the `sphcover ~ habitat` linear model to the `cover` dataset. See the summary of the model results with the `summary` function and the confidence intervals of the coefficients with the `confint` function. What is the interpretation of each coefficient? Are the confidence intervals plausible?

**Solution**

```{r}
mod <- lm(sphcover ~ habitat, data = cover)
summary(mod)
```

The intercept is the mean cover for the reference group (drained swamps, *Dr*). The coefficient `habitatRe` (respectively, `habitatUn`) is the difference between the mean cover of rewetted and drained swamps (respectively, the difference between undrained and drained swamps).

```{r}
confint(mod)
```

The confidence interval for the mean of the *Dr* group reaches negative values, which is not possible for a percentage cover.


c) Create a function with arguments `x` and `i`, which fits the linear model in b) by replacing the original data set (`data = cover`) with `data = x[i, ]`, and then returns the coefficients of the model with the function `coef`. Then, apply `boot` to the `cover` dataset with the function created and perform 10,000 replicates.

*Notes*

- When the first argument of `boot' is a dataset, it is the rows of that dataset that are resampled.

- Since the statistic calculated by the function has several values (each of the coefficients), the `t` element of the `boot` result is a matrix rather than a vector. The columns of this matrix correspond to each of the coefficients in order. You can calculate a statistic for each column with the function `apply`, e.g. `apply(res$t, 2, mean)`. Here, `2` indicates to calculate the `mean` function per column (`1` would mean per row).

**Solution**

```{r}
lm_hab <- function(x, i)  {
    mod <- lm(sphcover ~ habitat, x[i,])
    coef(mod)
}

boot_hab <- boot(cover, lm_hab, R = 10000)

# Mean
apply(boot_hab$t, 2, mean)

# Standard error
apply(boot_hab$t, 2, sd)
```

The mean of the bootstrap is very close to the coefficients estimated in b), but the standard errors have decreased. 

Note that all coefficients depend on the mean of the reference group *Dr*. The mean and standard deviation of this group are strongly affected by two extreme values around 20, as shown in graph a). These extreme values will be absent in several bootstrap samples, which has the effect of reducing the estimated standard deviation for the *Dr* group and thus reducing the standard error of the model's coefficients.


d) The application of the bootstrap in c) resamples across all rows, so that the number of observations in each habitat type varies from sample to sample. If it is preferable to consider these numbers as fixed quantities, habitat types can be defined as strata by adding the argument `strata = as.factor(cover$habitat)` to the `boot` function. (The conversion of the variable `habitat` to a factor is necessary here.)

Repeat the analysis in c) with stratified resampling and compare the standard errors obtained for each coefficient.

**Solution**

```{r}
boot_strat <- boot(cover, lm_hab, R = 10000, strata = as.factor(cover$habitat))

# Mean
apply(boot_strat$t, 2, mean)

# Standard error
apply(boot_strat$t, 2, sd)
```

All standard errors have decreased compared to c).


e) Calculate the confidence interval for the `habitatUn` coefficient according to the result of the bootstrap in d). Note that you need to add the argument `index = 3` to the function `boot.ci` to tell R to calculate the interval for the 3rd coefficient.

**Solution**

```{r}
boot.ci(boot_strat, index = 3)
```


f) Finally, we will resample the model residuals. 

- Fit a linear model as in b), then add to the dataset `cover` a column for the model's `fitted` values.  

- Write a function that creates a new dataset by adding a resampled vector `x[i]` to the expected values to produce a new response variable, then fits a model with this new response variable to the habitat.

- Simulate 10,000 samples with the `boot` function, using as arguments (1) the residuals vector (`residuals`) of the model and (2) the function created above. Do not specify strata. Recalculate the mean, standard error and 95% confidence interval of the coefficients.

Is resampling the residuals a good choice for these data?

**Solution**

```{r}

mod <- lm(sphcover ~ habitat, data = cover)
cover$fitted <- fitted(mod)

lm_resid <- function(x, i) {
    cover$cover_new <- cover$fitted + x[i]
    mod_new <- lm(cover_new ~ habitat, data = cover)
    coef(mod_new)
}

boot_resid <- boot(residuals(mod), lm_resid, R = 10000)

# Mean
apply(boot_resid$t, 2, mean)

# Standard error
apply(boot_resid$t, 2, sd)

boot.ci(boot_resid, index = 3)
```

Resampling of the residuals assumes that the residuals in each group are from the same distribution. In this case, the variances are not homogeneous between groups, so it is preferable to re-sample the observations.
