---
title: "Randomization tests - Solutions"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The file [sablefish.csv](../donnees/sablefish.csv) contains data from Kimura (1988) on the number of sablefish caught per unit effort (*catch*) in four Alaskan locations for each of the six years between 1978 and 1983.

```{r}
sable <- read.csv("../donnees/sablefish.csv")
head(sable)
```

a) Fit a linear model of catch as a function of location only. What is the interpretation of the `locationYakutat` coefficient of this model?

**Solution**

```{r}
lm_sable <- lm(catch ~ location, sable)
summary(lm_sable)
```

The coefficient `locationYakutat` gives the mean difference in catch between Yakutat and the reference location (Chirikof).

b) Perform a permutation test to calculate the $p$-value corresponding to the mean difference in catch between the Kodiak and Chirikof locations. Is this value consistent with the corresponding value in the linear model?

**Solution**

```{r}
set.seed(8202)

diff_perm <- function() {
   sable_perm <- sable
   sable_perm$location <- sample(sable$location) 
   mean(sable_perm$catch[sable_perm$location == "Kodiak"]) -
       mean(sable_perm$catch[sable_perm$location == "Chirikof"])
}

nperm <- 9999

diff_null <- replicate(nperm, diff_perm())

diff_obs <- mean(sable$catch[sable$location == "Kodiak"]) -
            mean(sable$catch[sable$location == "Chirikof"])

(sum(abs(diff_null) >= abs(diff_obs)) + 1) / (nperm + 1)
```

Yes, the value is close to the $p$-value for the `locationKodiak` coefficient of the model in a).

c) Using the *permuco* package, determine the $p$-value for the same difference, for a model including the additive effects of year and location. *Note*: We consider the year as a categorical variable here, so it must be converted to a factor. Does the $p$-value differ between the permutation test and the parametric model?

**Solution**

```{r}
library(permuco)
sable$year <- as.factor(sable$year)
lmperm(catch ~ year + location, sable)
```

The $p$-value is about 0.08 for both the parametric model and the permutation test.