---
title: "Introduction to Bayesian analysis - Solutions"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data

Already used for maximum likelihood exercises, the [thermal_range.csv](../donnees/thermal_range.csv) dataset represents the result of an experiment to determine the effect of temperature (*temp*) on the number of eggs (*num_eggs*) produced by a species of mosquito. Three replicates were measured for temperature values between 10 and 32 degrees Celsius.

```{r, warning = FALSE, message = FALSE}
library(brms)

therm <- read.csv("../donnees/thermal_range.csv")
head(therm)
```


## Bayesian estimation of the thermal optimum model

Let's remember the model used previously for this dataset. The average number of eggs produced is given by a Gaussian curve:

$$N = N_o \exp \left( - \frac{(T - T_o)^2}{\sigma_T^2} \right)$$

In this equation, $T_o$ is the optimum temperature, $N_o$ is the number of eggs produced at this optimum and $\sigma_T$ represents the tolerance around the optimum (the higher $\sigma_T$ is, the slower $N$ decreases around the optimum).

a) It is possible to estimate the parameters of a non-linear model like this one in *brms*. For example:

```{r, eval = FALSE}
brm(bf(num_eggs ~ No * exp(-(temp-To)^2/sigmaT^2), No + To + sigmaT ~ 1, nl = TRUE),
    data = therm)
```

*Note*:

- We need to enclose the formula in a `bf` function and specify the argument `nl = TRUE` (for non-linear).

- After the non-linear formula of the model, we need to add a term describing the parameters. Here, `No + To + sigmaT ~ 1` only means that we estimate a fixed effect for each parameter. If one of the parameters varied according to a group variable, we could write for example `No ~ (1|group), To + sigmaT ~ 1`. 

Since we are going to use a negative binomial distribution with a logarithmic relationship to represent the mean of the response (`family = negbinomial` in *brms*), we need to modify the formula above to represent the logarithm of the mean number of eggs $N$. Rewrite the `bf` function by applying this transformation.

**Solution**

$$\log N = \log N_o - \frac{(T - T_o)^2}{\sigma_T^2}$$

```{r, eval = FALSE}
brm(bf(num_eggs ~ logNo - (temp-To)^2/sigmaT^2, logNo + To + sigmaT ~ 1, nl = TRUE),
    data = therm, family = negbinomial)
```


b) Choose appropriate prior distributions for three parameters in the equation obtained above. In the `set_prior` statement, the parameter name is specified with `nlpar` for a non-linear model. For example, `set_prior("normal(0, 1)", nlpar = "To")` assigns a standard normal distribution to the parameter `To`. 

*Note*: Don't forget to specify the lower bound for `sigmaT`.

Also add a prior distribution for the $\theta$ parameter of the negative binomial distribution with `set_prior("gamma(2, 0.1)", class = "shape")`. You can visualize this distribution in R with `plot(density(rgamma(1E5, 2, 0.1))`. Since the variance of the negative binomial distribution is $\mu + \mu^2/\theta$, where $\mu$ is the mean, we want to avoid values of $\theta$ too close to zero. With the specified parameters, $\theta$ is small for values close to 0 and greater than 50 (with a $\theta$ so large, the negative binomial distribution almost matches that of Poisson).

**Solution**

Here is one possible prior choice:

```{r}
prior_therm <- c(set_prior("normal(4, 2)", nlpar = "logNo"),
                 set_prior("normal(20, 10)", nlpar = "To"),
                 set_prior("normal(0, 5)", nlpar = "sigmaT", lb = 0),
                 set_prior("gamma(2, 1)", class = "shape"))

```

- The `normal(4, 2)` distribution for $\log N_o$ gives ~95% probability to values of $\log N_0$ between 0 and 8, so $N_0$ between 1 and 3000 approximately.

- The `normal(20, 10)` distribution for $T_o$ gives ~95% probability to values between 0 and 40 degrees C.

- The half-normal distribution (normal truncated at 0) for $\sigma_T$ gives ~95% probability to values below 10.

Note that considering the temperature range tested in this experiment (between 10 and 32 degrees C), we could not detect an optimum beyond this range anyway, or a standard deviation that would be much greater than the difference between the extreme values tested.


c) Fit the non-linear model with `brm`, using the formula and prior distributions specified in the previous parts, with a negative binomial distribution of the response. Visualize the shape of the estimated $N$ vs. $T$ function with `marginal_effects`. Determine the mean value and the 95% credibility interval for the posterior distribution of each parameter.

**Solution**

```{r, eval = FALSE}
therm_fit <- brm(bf(num_eggs ~ logNo - (temp-To)^2/sigmaT^2, 
                    logNo + To + sigmaT ~ 1, nl = TRUE),
                 data = therm, family = negbinomial, prior = prior_therm)
```

```{r, cache = TRUE, include = FALSE}
therm_fit <- brm(bf(num_eggs ~ logNo - (temp-To)^2/sigmaT^2, 
                    logNo + To + sigmaT ~ 1, nl = TRUE),
                 data = therm, family = negbinomial, prior = prior_therm)
```

```{r}
marginal_effects(therm_fit)
```

```{r}
summary(therm_fit)
```


d) Compare the results in (c) with the maximum likelihood estimates and confidence intervals obtained in lab 3, reproduced in the table below.

Parameter | Estimate | Interval
----------|--------|-----------
$N_o$ | 123.2 | (104.2, 147.2)
$T_o$ | 23.9 | (23.4, 24.5)
$sigma_T$ | 6.82 | (6.33, 7.42)
$k$ | 0.103 | (0.059, 0.186)

*Note*: The parameter $k$ corresponds to $1 / \theta$ for the negative binomial distribution.

**Solution**

The results for $T_o$ and $\sigma_T$ are very close (considering the margin of error) for both methods.

For the other two parameters `logNo` and `shape`, we can transform the bounds of the intervals to compare with $N_o$ and $k$ (but since $k$ is the inverse of `shape`, we must swap the bounds). We can also transform the posterior mean, but we do not expect it to be equal to the maximum likelihood estimate. 

```{r}
exp(c(4.82, 4.61, 5.04)) # N_o = exp(log_No)
1/(c(6.09, 9.39, 3.50)) # k = 1/shape
```

The estimate of $N_o$ is consistent for both methods, but the $k$ parameter is higher for the Bayesian method (meaning more overdispersion) compared to the maximum likelihood estimate. 


e) Check the posterior prediction intervals with `pp_check(..., type = "intervals")`. Do the observations appear to be consistent with the fitted model?

**Solution**

```{r}
pp_check(therm_fit, type = "intervals")
```

If the model is good, we expect that about 50% of the points will be within the interval in bold and 90% within the thinner line, which seems to be the case here.

f) As we will see next week, the *Stan* program that *brms* uses produces a sample of the joint posterior distribution of the model parameters. By default, this sample includes 4000 parameter sets. The `posterior_epred` function of *brms* calculates the mean prediction according to each of these parameter sets for a new dataset given by the `newdata` argument, like the `predict` function in the case of regression models.

Use the `posterior_epred` function to calculate the ratio of mean egg production at 25 degrees C compared to 20 degrees C, and a 95% credibility interval for this ratio.

**Solution**

We first call `posterior_epred` with a dataset containing the two desired temperature values.

```{r}
post_pred <- posterior_epred(therm_fit, newdata = data.frame(temp = c(20, 25)))
str(post_pred)
```

The result is a matrix of 4000 rows (one per parameter set from the posterior distribution) and 2 columns (for $T$ = 20 C and $T$ = 25 C).

The question asks for the mean and 95% credibility interval for the ratio between the means for these two temperatures, i.e. each value in column 2 divided by the corresponding value in column 1.

```{r}
N_20_25 <- post_pred[, 2] / post_pred[ ,1]
mean(N_20_25)
quantile(N_20_25, probs = c(0.025, 0.975))
```




