---
title: "Time series"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

The [oak_seeds.csv](../donnees/oak_seeds.csv) dataset shows the number of seeds of *Quercus crispula* oak collected annually (1980-2017) by 16 traps located in a stand of this species in Japan.

```{r}
seed <- read.csv("../donnees/oak_seeds.csv")
head(seed)
```

The [oak_weather.csv](../donnees/oak_weather.csv) file contains annual weather data for that same site:

```{r}
weather <- read.csv("../donnees/oak_weather.csv")
head(weather)
```

- *temp_fl*: Mean temperature ($^{\circ}$C) during the flowering period of the tree.
- *temp_gr*: Mean temperature ($^{\circ}$C) during the growing season.
- *rain_fl*: Total amount of rain (in mm) during the flowering period of the tree.
- *rain_gr*: Total amount of rain (in mm) during the growing season.

These data come from the following study:

> Shibata, M., Masaki, T., Yagihashi, T., Shimada, T., & Saitoh, T. (2019). Data from: Decadal changes in masting behaviour of oak trees with rising temperature. Dryad Digital Repository. [https://doi.org/10.5061/dryad.v6wwpzgrb](https://doi.org/10.5061/dryad.v6wwpzgrb)


## 1. Site-level time series

(a) Calculate the total number of seeds collected per year (all traps combined) and apply a square root transformation to the result. Convert the result into a temporal data frame (*tsibble*) and view the resulting time series.

*Note*: Since we will be using linear rather than generalized models in this exercise, the square root transformation is intended to stabilize the variance of the count data.

(b) Visualize the temporal correlations for this series. What type of ARIMA model (AR and/or MA, as well as their order) might be appropriate here?

(c) Fit an ARIMA model by letting the function automatically choose the type and order of the model. What do the estimated coefficients mean?

(d) Join the `weather` dataset and fit an ARIMA model with the four weather variables as external predictors. Do these variables help to better predict the number of seeds produced per year?

(e) What type of model is chosen by `ARIMA()` if you consider only the sub-series starting in the year 2000, without an external predictor? Explain this choice from the graph in (a) and the temporal correlations for this subseries.

(f) Calculate the forecasts from the models in (c) and (e) for the next five years. How do these forecasts differ?


## 2. Trap-level time series

(a) Go back to the original table showing the number of seeds per year and trap, then apply the square root transformation to the number of seeds. Then use the `lme` function from the *nlme* package to fit a linear mixed model including: the fixed effect of weather variables, the random effect of the trap and the temporal correlations from one year to another. 

Here is an example of how to specify a random effect of a `GROUP` variable on the intercept of a `lme` model, as well as an ARMA correlation between successive elements of the same `GROUP`:

```{r, eval = FALSE}
library(nlme)
mod_lme <- lme(..., data = ...,
             random = list(GROUP = ~1), 
             correlation = corARMA(p = ..., q = ..., form = ~ 1 | GROUP))
```

(b) Compare the accuracy of the fixed effects in this model in (a) to the model in 1(d). What is the reason for this difference?


