---
title: "Time series - Solutions"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

The [oak_seeds.csv](../donnees/oak_seeds.csv) dataset shows the number of seeds of *Quercus crispula* oak collected annually (1980-2017) by 16 traps located in a stand of this species in Japan.

```{r}
seed <- read.csv("../donnees/oak_seeds.csv")
head(seed)
```

The [oak_weather.csv](../donnees/oak_weather.csv) file contains annual weather data for that same site:

```{r}
weather <- read.csv("../donnees/oak_weather.csv")
head(weather)
```

- *temp_fl*: Mean temperature ($^{\circ}$C) during the flowering period of the tree.
- *temp_gr*: Mean temperature ($^{\circ}$C) during the growing season.
- *rain_fl*: Total amount of rain (in mm) during the flowering period of the tree.
- *rain_gr*: Total amount of rain (in mm) during the growing season.

These data come from the following study:

> Shibata, M., Masaki, T., Yagihashi, T., Shimada, T., & Saitoh, T. (2019). Data from: Decadal changes in masting behaviour of oak trees with rising temperature. Dryad Digital Repository. [https://doi.org/10.5061/dryad.v6wwpzgrb](https://doi.org/10.5061/dryad.v6wwpzgrb)


## 1. Site-level time series

(a) Calculate the total number of seeds collected per year (all traps combined) and apply a square root transformation to the result. Convert the result into a temporal data frame (*tsibble*) and view the resulting time series.

*Note*: Since we will be using linear rather than generalized models in this exercise, the square root transformation is intended to stabilize the variance of the count data.

**Solution**

```{r, message = FALSE, warning = FALSE}
library(fpp3)
library(dplyr)
library(ggplot2)

seed <- group_by(seed, year) %>%
    summarize(seeds = sum(seeds)) %>%
    mutate(sqrt_seeds = sqrt(seeds))

seed <- as_tsibble(seed, index = year)
autoplot(seed, sqrt_seeds)
```

(b) Visualize the temporal correlations for this series. What type of ARIMA model (AR and/or MA, as well as their order) might be appropriate here?

**Solution**

```{r}
cowplot::plot_grid(autoplot(ACF(seed, sqrt_seeds)), 
                   autoplot(PACF(seed, sqrt_seeds)))
```

The ACF and PACF are significant for a lag of 1, suggesting an MA(1) or AR(1) model.

(c) Fit an ARIMA model by letting the function automatically choose the type and order of the model. What do the estimated coefficients mean?

**Solution**

```{r}
mod_1c <- model(seed, ARIMA(sqrt_seeds))
report(mod_1c)
```

This is an MA(1) model. The coefficient MA1 is -0.69, which means that the residual in one year contributes negatively to the following year. The term `constant` (14.6) is the mean of `sqrt_seeds` across years.

(d) Join the `weather` dataset and fit an ARIMA model with the four weather variables as external predictors. Do these variables help to better predict the number of seeds produced per year?

**Solution**

```{r}
seed <- inner_join(seed, weather)

mod_1d <- model(seed, ARIMA(sqrt_seeds ~ temp_gr + rain_gr + temp_fl + rain_fl))
report(mod_1d)
```

Only the temperature during the flowering period (`temp_fl`) seems to have a significant effect, but the AIC is higher than the previous model, so this model is not better than the one without predictors.

(e) What type of model is chosen by `ARIMA()` if you consider only the sub-series starting in the year 2000, without an external predictor? Explain this choice from the graph in (a) and the temporal correlations for this subseries.

**Solution**

```{r}
seed2000 <- filter(seed, year >= 2000)
mod_1e <- model(seed2000, ARIMA(sqrt_seeds))
report(mod_1e)
```

It is now an AR(1) model with a negative correlation of -0.77. From the year 2000, the time series in 1(a) seems to alternate more regularly between high and low values every 2 years. Also, if the PACF graph contains a single significant value, the ACF is significantly positive at a two-year lag, as is the case for an AR model. (The negative correlation between $y(t)$ and $y(t-1)$, then $y(t-1)$ and $y(t-2)$, causes a positive correlation between $y(t)$ and $y(t-2)$).

```{r}
cowplot::plot_grid(autoplot(ACF(seed2000, sqrt_seeds)),
                   autoplot(PACF(seed2000, sqrt_seeds)))
```

(f) Calculate the forecasts from the models in (c) and (e) for the next five years. How do these forecasts differ?

```{r}
cowplot::plot_grid(autoplot(forecast(mod_1c, h = 5), seed),
                   autoplot(forecast(mod_1e, h = 5), seed))
```

In the 2nd year, the MA(1) model loses all "memory" and returns to the series mean, while the AR(1) model continues a cycle every 2 years.

## 2. Trap-level time series

(a) Go back to the original table showing the number of seeds per year and trap, then apply the square root transformation to the number of seeds. Then use the `lme` function from the *nlme* package to fit a linear mixed model including: the fixed effect of weather variables, the random effect of the trap and the temporal correlations from one year to another. 

Here is an example of how to specify a random effect of a `GROUP` variable on the intercept of a `lme` model, as well as an ARMA correlation between successive elements of the same `GROUP`:

```{r, eval = FALSE}
library(nlme)
mod_lme <- lme(..., data = ...,
             random = list(GROUP = ~1), 
             correlation = corARMA(p = ..., q = ..., form = ~ 1 | GROUP))
```

**Solution**

```{r}
seed <- read.csv("../donnees/oak_seeds.csv")
seed <- mutate(seed, sqrt_seeds = sqrt(seeds))
seed <- inner_join(seed, weather)
```

We try a MA(1) model as in 1(d).

```{r, message = FALSE, warning = FALSE}
library(nlme)
mod_lme <- lme(sqrt_seeds ~ temp_gr + rain_gr + temp_fl + rain_fl, 
               data = seed, random = list(trap = ~1), 
               correlation = corARMA(p = 0, q = 1, form = ~ 1 | trap))

summary(mod_lme)
```


(b) Compare the accuracy of the fixed effects in this model in (a) to the model in 1(d). What is the reason for this difference?

**Solution**

Here, three of the climate variables have a significant effect, but the MA(1) effect is also smaller (Theta1 = -0.26). 

The square root transformation applied to the number of seeds in each trap is not equivalent to the square root transformation applied to the sum of seeds in all traps. (The square root of the sum is not equal to the sum of the square roots). It appears that the residual variance is smaller in this model than in the model for the square root of the sum, which increases the accuracy of the fixed effect estimates.


