---
title: "Spatial data"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

The [bryo_belg.csv](data/bryo_belg.csv) dataset is adapted from the data of this study:

> Neyens, T., Diggle, P.J., Faes, C., Beenaerts, N., Artois, T. et Giorgi, E. (2019) Mapping species richness using opportunistic samples: a case study on ground-floor bryophyte species richness in the Belgian province of Limburg. *Scientific Reports* 9, 19122. https://doi.org/10.1038/s41598-019-55593-x

This data frame shows the specific richness of ground bryophytes (*richness*) for different sampling points in the Belgian province of Limburg, with their position *(x, y)* in km, in addition to information on the proportion of forest (*forest*) and wetlands (*wetland*) in a 1 km$^2$ cell containing the sampling point.

```{r}
bryo_belg <- read.csv("../donnees/bryo_belg.csv")
head(bryo_belg)
```

## Fitting a geostatistical model

For this exercise, we will use the square root of the specific richness as the response variable. The square root transformation often allows to homogenize the variance of the count data in order to apply a linear regression. 

*Note*: To model the count data directly, for example with a Poisson distribution, and also include spatial dependence, we would need to use a custom likelihood model or a Bayesian model.

a) Fit a linear model of the transformed species richness to the proportion of forest and wetlands, without taking into account spatial correlations. What is the effect of the two predictors in this model?

**Solution**

```{r}
bryo_belg$sr <- sqrt(bryo_belg$richness)

bryo_lm <- lm(sr ~ forest + wetland, data = bryo_belg)
summary(bryo_lm)
```

There is a positive effect of the fraction of forests and a negative effect of the fraction of wetlands, both significant.

b) Calculate the empirical variogram of the model residuals in (a). Does there appear to be a spatial correlation between the points?

*Note*: The `cutoff` argument to the `variogram` function specifies the maximum distance at which the variogram is calculated. You can manually adjust this value to get a good view of the sill.

**Solution**

```{r, warning = FALSE, message = FALSE}
library(gstat)
vario <- variogram(sr ~ forest + wetland, locations = ~ x + y, data = bryo_belg,
                   cutoff = 50)
plot(vario)
```

Yes, the variance is smaller for short distances and seems to increase up to ~ 40 m

c) Re-fit the linear model in (a) with the `gls` function in the *nlme* package, trying different types of spatial correlations (exponential, Gaussian, spherical). Compare the models (including the one without spatial correlation) with the AIC.

**Solution**

```{r, message = FALSE, warning = FALSE}
library(nlme)

bryo_exp <- gls(sr ~ forest + wetland, data = bryo_belg,
                correlation = corExp(form = ~ x + y, nugget = TRUE))

bryo_gaus <- gls(sr ~ forest + wetland, data = bryo_belg,
                correlation = corGaus(form = ~ x + y, nugget = TRUE))

bryo_spher <- gls(sr ~ forest + wetland, data = bryo_belg,
                correlation = corSpher(form = ~ x + y, nugget = TRUE))

rbind(AIC(bryo_lm), AIC(bryo_exp), AIC(bryo_gaus), AIC(bryo_spher))
```

The spherical model has the smallest AIC, followed closely by the exponential model.

d) What is the effect of the proportion of forests and wetlands according to the model in (c)? Explain the differences between the conclusions of this model and the model in (a).

**Solution**

```{r}
summary(bryo_spher)
```

The magnitude of the coefficients is smaller, their standard error is larger, and the effect of wetlands is no longer significant. Because of the spatial correlations, our different points are not independent and part of the effect originally attributed to the predictor variables could be a coincidence due to the spatial correlations of both the response variable and the predictor variables. 
