---
title: "Generalized additive models"
output:
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Answers for this lab must be submitted on Moodle before March 25th at 5pm.

## Data

The file [portal_ot.csv](../donnees/portal_ot.csv) is a subset of the Portal database (used for labs 2 and 5) which contains the number of observed individuals (*n*) of the species *Onychomys torridus* (southern grasshopper mouse) in each plot for the years 1988 to 2002. Each plot was subjected to a treatment (*plot_type*) to exclude some or all rodents from the plot.

```{r}
portal_ot <- read.csv("../donnees/portal_ot.csv")
portal_ot$plot_type <- as.factor(portal_ot$plot_type)
portal_ot$plot_id <- as.factor(portal_ot$plot_id)
head(portal_ot)
```

*Note*: As indicated in the code above, the `plot_type` and `plot_id` categorical variables must be converted to factors before adjusting a GAM.


## 1. Estimating the overall population trend

For all questions in this section, you must fit a generalized additive model to estimate the demographic trend of the species taking into account the effect of treatments: `n ~ plot_type + s(year)`. For now, we will ignore the grouping of measurements in plots.

(a) First fit a GAM where the observations follow a Poisson distribution. Briefly describe how the number of individuals varies by year and treatment. Considering the linkage function used for this model, what does the additivity of the `plot_type' and `s(year)` effects mean?

(b) Is the default value of the number of basis functions $k$ sufficient to represent `s(year)` in the model in (a)? If necessary, refit the model with a higher $k$. What is the maximum value of $k$ you can use here?

(c) Is the data overdispersed with respect to your model?

(d) Fit a new GAM with the negative binomial distribution, specified as `family = "nb"` in the `gam` function. What is the estimate of the parameter $\theta$ for this model? Does the fit seem better than for the Poisson model? Are there still fit problems?


## 2. Adding a plot random effect

(a) From the negative binomial model in 1(d), include a random effect of plots on the intercept. Check the fit of the model, including the normality of the random effects and the presence or absence of overdispersion.

(b) Now fit a model with a random effect of plots on the mean demographic trend `s(year)`, using a term of type `bs = "fs"` as seen in the course. Compare this model to the model in (a) with the AIC. *Note*: The *AICcmodavg* package is not compatible with GAM, but you can calculate the AIC for each model with the `AIC` function.

(c) What is the fraction of deviance explained by the best model as determined in (b)?

(d) Finally, illustrate the estimated population trend for this species in each plot. To do so, add to the `portal_ot` dataset the values predicted by the best model in (b) and a 95% confidence interval, and then plot the observed data, the estimated temporal trends and their confidence intervals for each plot in the same graph.
